---
title: Exploratory Data Analysis on Red Wine Dataset to find which parameters affect
  the red wine quality
author: "Rahul Daware (rahuldaware19@gmail.com)"
output:
  html_document:
    fig_width : 12
    fig_height : 10
  pdf_document: default
---

### Step 1: Import Red Wine Dataset  
Let us read the CSV file into R and explain all 11 properties. These properties are explained below:  


1. **fixed.acidity**: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)  
2. **volatile.acidity**: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste  
3. **citric.acid**: found in small quantities, citric acid can add 'freshness' and flavor to wines  
4. **residual.sugar**: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet  
5. **chlorides**: the amount of salt in the wine  
6. **free.sulfur.dioxide**: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine  
7. **total.sulfur.dioxide**: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  
8. **density**: the density of water is close to that of water depending on the percent alcohol and sugar content  
9. **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale  
10. **sulphates**: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant  
11. **alcohol**: the percent alcohol content of the wine  
12. **quality** : score between 0 and 10. 10 being the best  

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
require(ggplot2)
require(cowplot)
require(GGally)
```

Let us look at the structure of dataset:

```{r echo=FALSE}
wineData = read.csv("wineQualityReds.csv")
str(wineData)
```

Next, let us look at data set summary:
```{r echo=FALSE}
summary(wineData)
```

### Step 2: Univariate Analysis  


The histogram of Red Wine Quality is drawn below : 
```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$quality)) + 
  scale_x_continuous("Wine Quality") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Quality") +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") 
```
From the above graph, we can see that the quality of red wine peaks at 5 and drops down on extreme ends. This almost follows normal distribution.   


Now let us look at some other variables before moving into advanced plots, let us plot histogram of acidities

```{r echo=FALSE, warning=FALSE}
p1 = ggplot(data = wineData, aes(x=wineData$fixed.acidity)) +  
    scale_x_continuous("Fixed Acidity") + 
    scale_y_continuous("Count") +
    ggtitle("Histogram of Fixed Acidity") +
    geom_histogram(binwidth=1, fill = "blue", color = "black") 
  
p2 = ggplot(data = wineData, aes(x=wineData$volatile.acidity)) +
     scale_x_continuous("Volatile Acidity") + 
     scale_y_continuous("Count") +
     ggtitle("Histogram of Volatile Acidity") +
  geom_histogram(binwidth=0.25, fill = "blue", color= "black")

plot_grid(p1, p2)
```

The above plots the histogram of fixed acidity and volatile acidity. The fixed acidity falls in the range of 6 to 10. The volatile acidit peaks at 0.5.


Also let us plot histograms with all other variables:  

Below is the histogram of Red Wine Residual Sugar

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$residual.sugar)) + 
  scale_x_continuous("Wine Residual Sugar") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Residual Sugar") +
  geom_histogram(binwidth = 0.25, fill = "blue", color = "black") 
```

Below is the histogram of Red Wine Chlorides

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$chlorides)) + 
  scale_x_continuous("Red Wine Chlorides") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Chlorides") +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black") 
```


Below is the histogram os Red wine Free Sulfur Dioxide

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$free.sulfur.dioxide)) + 
  scale_x_continuous("Red Wine Free Sulfur Dioxide") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Free Sulfur Dioxide") +
  geom_histogram(binwidth = 4, fill = "blue", color = "black") 
```

Below is the histogram os red wine total sulfur dioxide

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$total.sulfur.dioxide)) + 
  scale_x_continuous("Red Wine Total Sulfur Dioxide") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Total Sulfur Dioxide") +
  geom_histogram(binwidth = 15, fill = "blue", color = "black")
```

Below is the histogram of red wine density

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$density)) + 
  scale_x_continuous("Red Wine Density") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Density") +
  geom_histogram(binwidth = 0.001, fill = "blue", color = "black")
```

Below is the histogram of red wine sulphates

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$sulphates)) + 
  scale_x_continuous("Red Wine Sulphates") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Sulphates") +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black")
```

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$citric.acid)) + 
  scale_x_continuous("Red Wine Citric Acid") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Citric Acid") +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black")
```

The histogram of pH level is shown below:

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$pH)) + 
  scale_x_continuous("pH") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine pH Level") +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") 
```


Below is the histogram of Citric Acid

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$citric.acid)) + 
  scale_x_continuous("Citric Acid") + 
  scale_y_continuous("Count") +
  ggtitle("Histogram of Red Wine Citric Acid") +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black") 
```


The citric acid content falls in the range of 0 to 0.6.  

### Step 3 : Bivariate Analysis

Lets draw correlation matrix of each variable with other.
```{r echo=FALSE, warning=FALSE}
exclude <- c("X")
keep <- !(names(wineData) %in% exclude)
theme_set(theme_minimal(20))
set.seed(2222)
wine_subset <- wineData[,keep]
names(wine_subset)
ggpairs(wine_subset)
```


To find out which parameter affects the quality, let us correlate quality with all other parameters
```{r echo=FALSE}

cor(wineData[,keep], wineData$quality)
```

From this, we can see that the quality of wine has less or negative correlation with acidity, no correlation with sugar and positive correlation with alcohol.  

Let us draw some scatter plots. Let us understand the relation between some parameters:  

```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$citric.acid, y = wineData$volatile.acidity)) + 
  scale_x_continuous("Citric Acid") + 
  scale_y_continuous("Volatile Acidity") +
  ggtitle("Volatile Acidity vs Citric Acid") +
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)
```
From the above plot, we can see that as citric acid content increases, the volatile acidity decreases.  

Next, let us plot Fixed Acidity versus Citric Acid
```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$citric.acid, y = wineData$fixed.acidity)) + 
  scale_x_continuous("Citric Acid") + 
  scale_y_continuous("Fixed Acidity") +
  ggtitle("Fixed Acidity vs Citric Acid") +
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)
```
From the above plot, we can see as citric acid content increases, fixed acidity increases.  

Let us try citric acid versus pH
```{r echo=FALSE, warning=FALSE}
ggplot(data = wineData, aes(x = wineData$citric.acid, y = wineData$pH)) + 
  scale_x_continuous("Citric Acid") + 
  scale_y_continuous("pH") +
  ggtitle("pH vs Citric Acid") +
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)
```
The above analysis looks correct. As citric acid content increases, pH decreases. This is evident from the correlation value between citric acid and pH. Correlation between quality and pH is -0.05 and correlation between quality and citric acid is about 0.22. This makes it clear that citric acid and pH are inversely proportional to each other.




Let us draw some scatter plots between two variables :


```{r echo=FALSE, warning=FALSE}
ggplot(data= wineData, 
       aes(y = wineData$quality, x = wineData$alcohol)) +
  scale_y_continuous("Quality") +
  scale_x_continuous("Alcohol") +
  ggtitle("Relationship between Quality and Alcohol") +
  geom_point()
```

The above plot clearly shows that as alcohol content increases, the quality increases.

```{r echo=FALSE, warning=FALSE}
ggplot(data= wineData, 
       aes(x = wineData$alcohol, y = wineData$citric.acid)) +
  scale_x_continuous("Alcohol") +
  scale_y_continuous("Citric Acid") +
  ggtitle("Relationship between Alcohol and Citric Acid") +
  geom_point()
```
 The above plot does not provide any clear idea of what is going on.

### Step 4 : Multivariate Analysis

Let us find the effect of alcohol and citric acid on quality of red wine.  

```{r echo=FALSE, warning=FALSE}
Quality <- factor(wineData$quality)
ggplot(data= wineData, 
       aes(x = wineData$alcohol, y = wineData$citric.acid, color = Quality)) +
  scale_x_continuous("Alcohol") +
  scale_y_continuous("Citric Acid") +
  ggtitle("Relationship between Quality, Alcohol and Citric Acid") +
  geom_point() +
  scale_colour_hue()
```

From the above plot, we can see that the quality is higher when both alcohol and citric acid content is higher.

Let us plot a few more plots and find out which parameters affect the quality.  

```{r set-options, echo=FALSE, cache=FALSE}
Quality <- factor(wineData$quality)
options(width = 500)
ggplot(data= wineData, 
       aes(x = wineData$alcohol, y = wineData$sulphates, color = Quality)) +
  scale_y_continuous("Sulphates") +
  scale_x_continuous("Alcohol") +
  ggtitle("Relationship between Quality, Alcohol and Sulphates") +
  geom_point() +
  scale_colour_hue()
```

Here we see that the sulphates show very less effect on the quality of red wine.

Let us take one more example. Let us find if volatile acidity and alcohol have any effect on quality.
```{r echo=FALSE, cache=FALSE}
Quality <- factor(wineData$quality)
ggplot(data= wineData, 
       aes(x = wineData$alcohol, 
           y = wineData$volatile.acidity, color = Quality)) +
  scale_y_continuous("Volatile Acidity") +
  scale_x_continuous("Alcohol") +
  ggtitle("Relationship between Quality, Alcohol and Volatile Acidity") +
  geom_point() +
  scale_colour_hue()
```

This plot gives a guess that higher the volatile acidity, lower is the quality. But this cannot be confirmed for sure.

Next, let us find if pH and alcohol have any effect on quality.

```{r echo=FALSE, cache=FALSE}
Quality <- factor(wineData$quality)
ggplot(data= wineData, 
       aes(x = wineData$alcohol, y = wineData$pH, color = Quality)) +
  scale_y_continuous("pH") +
  scale_x_continuous("Alcohol") +
  ggtitle("Relationship between Quality, Alcohol and pH") +
  geom_point() +
  scale_colour_hue()
```

From the above plot, it is difficult to make any conclusion.  

Finally, let us group by four range of quality. Let us call it bad, average, good and very good.   
Bad is when quality is between 0-3, average between 3-5, good between 5-8 and very good is 8-10. All of them are not including upper bound.

```{r echo=FALSE, cache=FALSE}
wineData$category[wineData$quality < 4] = "bad"
wineData$category[wineData$quality >= 4 & wineData$quality < 6] = "average"
wineData$category[wineData$quality >= 6 & wineData$quality < 8] = "good"
wineData$category[wineData$quality >= 8] = "very.good"
```
Let us use this category value and plot the above three plots again. First let us plot the histogram of category.

```{r echo=FALSE, cache=FALSE}
wineData$category <- factor(wineData$category, 
                            levels = c("bad","average","good","very.good"))
ggplot(data = wineData, aes(x = wineData$category)) + 
  scale_x_discrete("Category") + 
  scale_y_discrete("Count") +
  ggtitle("Histogram of Red Wine Category") +
  geom_bar(fill = "blue", color = "black", stat = "count") 
```

Now, let us plot the above multivariate graphs with the category variable. This will give more clear information.

```{r echo=FALSE, warning=FALSE}
ggplot(data= wineData, aes(x = wineData$alcohol, 
                           y = wineData$citric.acid, color = category)) +
  scale_y_continuous("Citric Acid") +
  scale_x_continuous("Alcohol") +
  ggtitle("Relationship between Category, Alcohol and Citric Acid") +
  geom_point()
```

### Step 5: Final Plots

Now, let us plot the above multivariate graphs with the category variable. This will give more clear information. The above graph looks very informative. Let us use that plot and change its color scheme to make it more visible.

```{r echo=FALSE, warning=FALSE}
ggplot(data= wineData, aes(x = wineData$alcohol, y = wineData$citric.acid, color = category)) +
  scale_y_continuous("Citric Acid (g / dm^3)") +
  scale_x_continuous("Alcohol % by volume") +
  ggtitle("Relationship between Category, Alcohol and Citric Acid") +
  geom_point() +
  scale_colour_brewer(palette = "Set1")
```
Above plot tends to show that higher the alcohol, higher is the quality of wine. The citric acid content does not clearly show any significant impact on the quality.  

```{r echo=FALSE, cache=FALSE}
ggplot(data= wineData, aes(x = wineData$alcohol, 
                           y = wineData$sulphates, color = category)) +
  scale_y_continuous("Sulphates (g / dm3)") +
  scale_x_continuous("Alcohol (% by volume)") +
  ggtitle("Relationship between Category, Alcohol and Sulphates") +
  geom_point() +
  scale_colour_brewer(palette = "Set1")
```

From the above plot, we can say with good confidence that sulphate content above 0.75 will have a better quality than the one less than 0.75.  


```{r echo=FALSE, cache=FALSE}
ggplot(data= wineData, aes(x = wineData$alcohol, y = wineData$volatile.acidity, color = category)) +
  scale_y_continuous("Volatile Acidity (g / dm^3)") +
  scale_x_continuous("Alcohol (% by volume)") +
  ggtitle("Relationship between Category, Alcohol and Volatile Acidity") +
  geom_point() +
  scale_colour_brewer(palette = "Set1")
```

From the above plot, we can see that higher the volatile acidity, lower is the quality of wine. This is also verified from the correlation values.

### Step 6: Reflection

It is time to reflect all the observations. First, we tried to observe how each individual variable varies. We found that quality of red wine follows normal distribution with majority of quality values fall near to the mean. The fixed acidity and volatile acidity also follows normal distribution. The residual sugar value does not seem to follow normal ditribution. Most values fall near the range of 1 to 3 while some outliers extending upto 15. The chlorides have most values less than 0.15. Some values fall in the higher range. It would be interesting to see if any of these outliers have higher quality. The Free Sulfur Dioxide value and Total Sulfur Dioxide value also fall in the lower range. The red wine density seems to follow normal distribution with mean at around 0.9975.  

In the bivariate analysis, we find the correlation of quality with rest other variables. We see that Alcohol content has the highest correlation with quality followed by volatile acidity and sulphates. The correlation values also show that residual sugar and pH should have very less effect on the quality as these correlation values are closer to 0. We also saw two plots. The alcohol versus quality plot also clearly showed that quality increases as alcohol content increases.   

In the multivariate analysis, the plots did not give a very clear idea. But we could see that the alcohol content has positive effect on the quality. To get a clear picture, we binned the quality into four categories (bad, average, good, very good). Most of the values fell into average and good bins. We again ran the plots with these values and found out that alcohol clearly has positive effect on the quality. When alcohol and citric acid both are high, there is higher probability for the red wine to be of good quality. The suphates did not show much variation. They had very less effect on the quality. The volatile acidity clearly showed negative effect on the quality. Lower the volatile acidity, higher would be the quality.  

From personal point of view, there could be a lot more possibilities as there are 11 variables to be considered. We can have multi dimensional analysis which would bring out the exact dependencies of quality of red wine. 

The further analysis of this project would include comparing every two variables against the quality and find which two variables affect the quality the most. We took alcohol in most of the example as the correlation showed positive influence of alcohol on acidity.



#### References :

The dataset was downloaded from the folowing reference :  



P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf  
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  

